---
title: "Analyses for the paper 'Changes in public perception of artificial intelligence in healthcare after exposure to ChatGPT'"
author: "Anders Aasted Isaksen"
format:
    html:
        embed-resources: true
---

## Load packages

```{r}
library(haven)
library(readr)
library(data.table)
library(lubridate)
library(here)
library(finalfit)
library(gridExtra)
library(nnet)
library(ggplot2)
library(ggalluvial)
library(dplyr)
library(WeightIt)
```

# Load data

```{r load}
#| include: false

# Load HICD data and label values from SAS formats and assign values from the labels:

# 2024 survey:
rmstenodiabetes2023_svar <-
    haven::read_sas(
        "E:/rawdata/707987/DST_survey_2024/rmstenodiabetes2023_svar.sas7bdat",
        "E:/rawdata/707987/DST_survey_2024/rmstenodiabetes2023_formats.sas7bcat"
    )


survey_2024 <-
    rmstenodiabetes2023_svar |> dplyr::mutate_if(sjlabelled::is_labelled, sjlabelled::as_label)

setDT(survey_2024)

# 2022 Survey (no sas formats file):
survey_2022 <-
    haven::read_sas("E:/rawdata/707987/Eksterne data/20220503/_20220426_hicd22.sas7bdat")
setDT(survey_2022)


# Achtung: 62 duplicate responses in 2022 from 31 individuals

# Take .SD[1] to only keep one response per person.
# Filter survey data and wrangle NAs in 2022 to account for q4=3 (haven't heard of AI))
# True missing in q10 are only those missing in q4
# q4==3 is a separate level of q10
survey_2022_clean <-
    survey_2022[, .SD[1], by = CPR][, .(
        pnr = CPR,
        heard_of_ai = factor(q4),
        ai_perception_2022_raw = addNA(factor(q10))
    )]

survey_2022_clean[, ai_perception_2022 := fifelse(
    is.na(heard_of_ai) |
        (ai_perception_2022_raw == "NA" &
             !is.na(heard_of_ai)),
    NA,
    ai_perception_2022_raw
)]

survey_2024_clean <-
    survey_2024[, .(
        pnr,
        ai_perception_2024 = aiq1,
        chatgpt_use = aiq2)]

levels(survey_2022_clean$ai_perception_2022) <-
    levels(survey_2024_clean$ai_perception_2024)

```

## Format response data

```{r merge}
#| include: false
responses <- merge(survey_2024_clean[!(is.na(ai_perception_2024) |
                                           is.na(chatgpt_use))],
                   survey_2022_clean[!is.na(ai_perception_2022), .(pnr, ai_perception_2022)],
                   by = "pnr",
                   all = FALSE)

# Format responses:
levels(responses$chatgpt_use) <- c("ChatGPT use", "ChatGPT use", "No use", "No use")
levels(responses$ai_perception_2022) <-
    c("Benefits", "Equal", "Risks", "Don't know", "Don't know")
levels(responses$ai_perception_2024) <-
    c("Benefits", "Equal", "Risks", "Don't know", "Don't know")



responses$ai_perception_2022 <- factor(responses$ai_perception_2022,
                                       levels = c("Don't know", "Benefits", "Equal", "Risks"))
responses$ai_perception_2024 <- factor(responses$ai_perception_2024,
                                       levels = c("Don't know", "Benefits", "Equal", "Risks"))

responses[, ai_perception_2022 := relevel(ai_perception_2022, ref = "Don't know")]
responses[, ai_perception_2024 := relevel(ai_perception_2024, ref = "Don't know")]


```

## Transform to long format for alluvial plots

```{r long-format}
#| include: false

## Long format:
responses_long <- responses |> dplyr::mutate(pnr = dplyr::row_number()) |> tidyr::pivot_longer(
    cols = tidyr::starts_with("ai_perception"),
    names_to = "survey_year",
    values_to = "ai_perception"
)

responses_long <-
    responses_long |> dplyr::mutate(survey_year = factor(
        stringr::str_extract(survey_year, "....$"),
        levels = c("2022", "2024")
    ))
```

## Function to add external covariates

```{r covariates}
#| include: false

add_covariates <-
    function(study_population,
             index_year,
             education_year) {
        # Load data:
        
        # Bef
        bef_data <-
            readRDS(
                paste0(
                    "E:/workdata/707987/aai_workspace/osdc2023/data/raw/bef",
                    index_year - 1,
                    "12.sas7bdat.rds"
                )
            )
        setDT(bef_data)
        names(bef_data) <- tolower(names(bef_data))
        bef_data <-
            bef_data[, .(pnr, age = alder, sex = as.factor(ifelse(koen ==
                                                                      '1', 'Men', 'Women')))]
        
        # Education:
        edu_length_years <- function(education_year) {
            edu_data <-
                haven::read_sas(
                    paste0(
                        "E:/rawdata/707987/Grunddata/Leverance_3/POPB/udda",
                        education_year,
                        "09.sas7bdat"
                    )
                )
            formats <-
                read_sas(
                    "E:/Formater/SAS formater i Danmarks Statistik/SAS_datasaet/Disced/n_audd_pria_l1lx_k.sas7bdat"
                )
            setDT(edu_data)
            setDT(formats)
            edu_length <-
                merge(edu_data[, .(pnr = PNR, edu_code = HFAUDD)], formats[, .(
                    edu_code = START,
                    education_years = cut(
                        as.numeric(AUDD_PRIA_L1LX_K) / 12,
                        breaks = c(0, 10, 15, Inf),
                        right = FALSE
                    )
                )], by = "edu_code", all.x = TRUE)
            return(edu_length[, .(pnr, education_years)])
        }
        
        education_data <- edu_length_years(education_year)
        
        # Diabetes
        diabetes_data <-
            readr::read_csv(
                "E:/workdata/707987/_common/data/csv/diabetes_onset_imputed_hicd_2023.csv",
                col_types = cols(
                    diabetes_type = col_factor(levels = c("1", "2")),
                    date_onset_diabetes_hicd = col_date(format = "%Y-%m-%d")
                )
            )
        setDT(diabetes_data)
        names(diabetes_data) <- tolower(names(diabetes_data))
        diabetes_data <- diabetes_data[, .(pnr, prevalent_diabetes = date_onset_diabetes_hicd < as.Date(paste0(index_year, "-01-01")), diabetes_type)]
        
        # standardize colnames:
        names(study_population) <- tolower(names(study_population))
        
        
        
        # Merge together:
        output <- Reduce(
            function(x, y)
                merge(x, y, by = 'pnr', all.x = TRUE),
            list(study_population, bef_data, education_data, diabetes_data)
        )
        
        output[, diabetes := factor(ifelse(
            is.na(prevalent_diabetes) |
                prevalent_diabetes == FALSE,
            "No diabetes",
            diabetes_type
        ))]
        return(output)
    }

responses_covariates <-
    add_covariates(
        study_population = responses,
        index_year = 2022,
        education_year = 2021
    )


responses_covariates[, chatgpt_use := relevel(chatgpt_use, ref = "No use")]

responses_covariates[, diabetes := relevel(diabetes, ref = "No diabetes")]
levels(responses_covariates$diabetes) <- c("No diabetes", "T1D", "T2D")
responses_covariates <- responses_covariates |> dplyr::mutate_all(sjlabelled::remove_label)

```



## Define function for regression analyses

Main analysis with dynamic reference group to estimate change from baseline perception

```{r dynamic-ref-levels-analysis}
#| include: false

dynamic_ref_regression_func <- function(input_data, weights = NULL) {
    outcome_levels <- levels(input_data$ai_perception_2024)
    
    ref_levels_model_list <-
        vector(mode = "list", length = 0)
    
    
    
    for (baseline_perception in levels(input_data$ai_perception_2022)) {
        data_strata <- input_data[ai_perception_2022 == baseline_perception]
        
        
        for (ref_level in outcome_levels) {
            data_strata$ai_perception_2024 <-
                relevel(data_strata$ai_perception_2024, ref = ref_level)
            
            ref_levels_model_list[[baseline_perception]][[ref_level]] <-
                broom::tidy(
                    multinom(
                        ai_perception_2024 ~ chatgpt_use + sex + I(age / 10) + diabetes + education_years,
                        data = data_strata,
                        if (is.null(weights)) {
                            weights = NULL
                        } else {
                            weights = get(weights)
                        }
                    ),
                    conf.int = TRUE,
                    exponentiate = TRUE
                )
            
        }
        
    }
    
    ### Select models with the same baseline strata and outcome reference level
    
    matching_strata_outcome <-
        vector(mode = "list", length = 0)
    
    for (baseline_strata in names(ref_levels_model_list)) {
        matching_strata_outcome[[baseline_strata]] <-
            ref_levels_model_list[[baseline_strata]][[baseline_strata]]
        
    }
    
    matching_strata_dt <-
        data.table::rbindlist(matching_strata_outcome, idcol = "baseline_perception")
    
    # Add reference category values:
    matching_strata_dt_ref <- rbind(
        matching_strata_dt,
        data.frame(
            baseline_perception = rep(outcome_levels, 3),
            y.level = rep(outcome_levels, 3),
            term = c("chatgpt_useChatGPT use", "diabetesT1D", "diabetesT2D"),
            estimate = rep(1, length(outcome_levels) * 3),
            conf.low = rep(1, length(outcome_levels) * 3),
            conf.high = rep(1, length(outcome_levels) * 3)
        ),
        fill = TRUE
    )
    
    # Add helper variables:
    matching_strata_dt_ref[, estimates_label := ifelse(
        estimate != 1,
        paste0(
            round_tidy(estimate, 2),
            " (",
            round_tidy(conf.low, 2),
            "-",
            round_tidy(conf.high, 2),
            ")"
        ),
        paste0("(ref)")
    )]
    
    # Remove facet_wrap headers:
    matching_strata_dt_ref[, baseline_perception_empty := factor(baseline_perception, labels = c(" ", " ", " ", " "))]
    
    return(matching_strata_dt_ref)
}

```

## Create baseline tables

```{r table-one}
#| include: false

# Exclude missing data:
responses_covariates <-
    responses_covariates[complete.cases(responses_covariates[, .(pnr,
                                                                 ai_perception_2022,
                                                                 ai_perception_2024,
                                                                 age,
                                                                 diabetes,
                                                                 education_years)])]

basetable <- responses_covariates |> summary_factorlist(
    dependent = "chatgpt_use",
    explanatory = c(
        "sex",
        "age",
        "diabetes",
        "education_years",
        "ai_perception_2022",
        "ai_perception_2024"
    ),
    column = TRUE,
    total_col = TRUE,
    add_col_totals = TRUE
    
)

basetable_by_baseline_perception <- responses_covariates |> summary_factorlist(
    dependent = "ai_perception_2022",
    explanatory = c(
        "sex",
        "age",
        "diabetes",
        "education_years",
        "ai_perception_2024",
        "chatgpt_use"
    ),
    column = TRUE,
    total_col = FALSE,
    add_col_totals = TRUE
    
)

summary_crosstable_chatgpt <- responses_covariates[chatgpt_use == "ChatGPT use"] |> summary_factorlist(
    dependent = "ai_perception_2022",
    explanatory = "ai_perception_2024",
    column = TRUE,
    total_col = FALSE,
    add_col_totals = TRUE
    
)

summary_crosstable_nonusers <- responses_covariates[chatgpt_use != "ChatGPT use"] |> summary_factorlist(
    dependent = "ai_perception_2022",
    explanatory = "ai_perception_2024",
    column = TRUE,
    total_col = FALSE,
    add_col_totals = TRUE
    
)
```

## Create alluvial plots

```{r}
# Work-around for not wanting the reference category to be the first level:

responses_covariates_alluvial <- responses_covariates
responses_covariates_alluvial$ai_perception_2022 <- factor(
    responses_covariates_alluvial$ai_perception_2022,
    levels = c("Don't know", "Benefits", "Equal", "Risks")
)
responses_covariates_alluvial$ai_perception_2024 <- factor(
    responses_covariates_alluvial$ai_perception_2024,
    levels = c("Don't know", "Benefits", "Equal", "Risks")
)

responses_long <- responses_covariates_alluvial |> dplyr::mutate(pnr = dplyr::row_number()) |> tidyr::pivot_longer(
    cols = tidyr::starts_with("ai_perception"),
    names_to = "survey_year",
    values_to = "ai_perception"
)

responses_long <-
    responses_long |> dplyr::mutate(survey_year = factor(
        stringr::str_extract(survey_year, "....$"),
        levels = c("2022", "2024")
    ))




alluv_users <-
    responses_long |> filter(chatgpt_use == "ChatGPT use") |> ggplot(aes(
        x = survey_year,
        stratum = ai_perception,
        alluvium = pnr,
        fill = ai_perception
    )) +
    geom_flow() +
    geom_stratum(alpha = 0.5) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank()) +
    MetBrewer::scale_fill_met_d("Hiroshige", direction = "-1") +
    labs(title = "ChatGPT exposed",
         fill = "Perception of AI",
         x = "",
         y = "Individuals") +
    theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        # axis.title.x = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold")
    )

alluv_nonusers <-
    responses_long |> filter(chatgpt_use != "ChatGPT use") |> ggplot(aes(
        x = survey_year,
        stratum = ai_perception,
        alluvium = pnr,
        fill = ai_perception
    )) +
    geom_flow() +
    geom_stratum(alpha = 0.5) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank()) +
    MetBrewer::scale_fill_met_d("Hiroshige", direction = "-1") +
    labs(title = "Unexposed",
         fill = "Perception of AI",
         x = "",
         y = "") +
    theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        # axis.title.x = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold")
    )

```

## Create weights

```{r make-weights}

# Weight study population to represent 2022 background population distributions of
# age, sex, diabetes, education

background_covariates <-
    add_covariates(
        study_population = as.data.table(
            readRDS(
                "E:/workdata/707987/aai_workspace/osdc2023/data/raw/bef202112.sas7bdat.rds"
            )
        )[, .(pnr)],
        index_year = 2022,
        education_year = 2021
    )

background_covariates[, in_study_pop := pnr %in% responses_covariates$pnr]
background_covariates <- background_covariates[, .(pnr, age, sex, education_years, diabetes, in_study_pop)]
background_covariates_targetpop <-
    background_covariates[complete.cases(background_covariates) &
                              age <= max(responses_covariates$age) &
                              age >= min(responses_covariates$age)]


# To create invitee-population weights: load the original invitee population in 2020:
invitees_2020 <- read_sas("E:/rawdata/707987/Fra DST Survey/stenormdiabetes2020_stik_.sas7bdat")$pnr

background_covariates_targetpop_invitees <- background_covariates_targetpop[pnr %in% invitees_2020]

# Compute regression models for weighting:
weightit_model_background <-
    WeightIt::weightit(
        in_study_pop ~ age + sex + diabetes + education_years,
        data = background_covariates_targetpop,
        method = "ps",
        stabilize = TRUE
    )

weightit_model_invitees <-
    WeightIt::weightit(
        in_study_pop ~ age + sex + diabetes + education_years,
        data = background_covariates_targetpop_invitees,
        method = "ps",
        stabilize = TRUE
    )

background_covariates_targetpop$weights_background <- weightit_model_background$weights

background_covariates_targetpop_invitees$weights_invitees <- weightit_model_invitees$weights


responses_weights <- Reduce(
    function(x, y)
        merge(x, y, by = 'pnr', all.x = TRUE),
    list(
        responses_covariates,
        background_covariates_targetpop[, .(pnr, weights_background)],
        background_covariates_targetpop_invitees[, .(pnr, weights_invitees)]
    )
)


# Cap weights:
responses_weights[, weights_background_capped := pmin(weights_background, quantile(weights_background, 0.95))]
responses_weights[, weights_invitees_capped := pmin(weights_invitees, quantile(weights_invitees, 0.95))]

# Rescale weights:
responses_weights[, weights_background_rescaled := weights_background_capped /
                      mean(weights_background_capped)]
responses_weights[, weights_invitees_rescaled := weights_invitees_capped /
                      mean(weights_invitees_capped)]

```

## Supplementary: unstratified regression analysis

```{r overall-analysis-weights}
#| eval: false

# This analysis is adjusted for baseline perception, rather than stratified by it
overall_analysis_weights <- broom::tidy(
    multinom(
        ai_perception_2024 ~ chatgpt_use + sex + I(age / 10) + diabetes + education_years + ai_perception_2022,
        data = responses_weights,
        weights = weights_invitees_rescaled
    ),
    conf.int = TRUE,
    exponentiate = TRUE
)
```

## Supplementary: weighted base table

```{r}
basetable_weighted <- responses_weights |> summary_factorlist(
dependent = "chatgpt_use",
explanatory = c(
"sex",
"age",
"diabetes",
"education_years",
"ai_perception_2022",
"ai_perception_2024"
),
column = TRUE,
total_col = TRUE,
add_col_totals = TRUE,
weights = "weights_background_rescaled"

)

basetable_weighted_invitees <- responses_weights |> summary_factorlist(
dependent = "chatgpt_use",
explanatory = c(
"sex",
"age",
"diabetes",
"education_years",
"ai_perception_2022",
"ai_perception_2024"
),
column = TRUE,
total_col = TRUE,
add_col_totals = TRUE,
weights = "weights_invitees_rescaled"
)

```



## Crosstables

```{r crosstables}

## Create weighted cross-tables:

responses_split <- split(responses_weights, responses_weights$chatgpt_use)

format_crosstable <- function(subset_chatgpt_use, weighted = FALSE) {
    # Counts & proportions
    
    if (weighted == TRUE) {
        counts <- round(
            xtabs(
                weights_background_rescaled ~ ai_perception_2024 + ai_perception_2022,
                data = subset_chatgpt_use
            )
        )
    } else {
        counts <- table(
            subset_chatgpt_use$ai_perception_2024,
            subset_chatgpt_use$ai_perception_2022
        )
    }
    
    
    props <- prop.table(counts, margin = 2) * 100 # Column percentages
    
    # Formatting output:
    formatted <- matrix(
        paste0(counts, " (", round(props, 1), "%)"),
        nrow = nrow(counts),
        ncol = ncol(counts),
        dimnames = dimnames(counts)
    )
    
    # Add column sums:
    col_sums <- colSums(counts)
    
    final_format <- rbind(Total = col_sums, formatted)
    return(final_format)
}

unweighted_crosstabs <- lapply(responses_split, weighted = FALSE, format_crosstable)

weighted_crosstabs <- lapply(responses_split, weighted = TRUE, format_crosstable)
```

## Supplementary: weighted alluvial plots

```{r}
## Long format:

# Work-around for not wanting the reference category to be the first level:

responses_weighted_alluvial <- responses_weights
responses_weighted_alluvial$ai_perception_2022 <- factor(
    responses_weighted_alluvial$ai_perception_2022,
    levels = c("Don't know", "Benefits", "Equal", "Risks")
)
responses_weighted_alluvial$ai_perception_2024 <- factor(
    responses_weighted_alluvial$ai_perception_2024,
    levels = c("Don't know", "Benefits", "Equal", "Risks")
)

weighted_long <- responses_weighted_alluvial |> dplyr::mutate(pnr = dplyr::row_number()) |> tidyr::pivot_longer(
    cols = tidyr::starts_with("ai_perception"),
    names_to = "survey_year",
    values_to = "ai_perception"
)

weighted_long <-
    weighted_long |> dplyr::mutate(survey_year = factor(
        stringr::str_extract(survey_year, "....$"),
        levels = c("2022", "2024")
    ))


weighted_alluv_users <-
    weighted_long |> filter(chatgpt_use == "ChatGPT use") |> ggplot(
        aes(
            x = survey_year,
            stratum = ai_perception,
            alluvium = pnr,
            y = weights_background_rescaled,
            fill = ai_perception
        )
    ) +
    geom_flow() +
    geom_stratum(alpha = 0.5) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank()) +
    MetBrewer::scale_fill_met_d("Hiroshige", direction = "-1") +
    labs(title = "ChatGPT exposed",
         fill = "Perception of AI",
         x = "",
         y = "Individuals") +
    theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        # axis.title.x = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold")
    )

weighted_alluv_nonusers <-
    weighted_long |> filter(chatgpt_use != "ChatGPT use") |> ggplot(
        aes(
            x = survey_year,
            stratum = ai_perception,
            alluvium = pnr,
            y = weights_background_rescaled,
            fill = ai_perception
        )
    ) +
    geom_flow() +
    geom_stratum(alpha = 0.5) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank()) +
    MetBrewer::scale_fill_met_d("Hiroshige", direction = "-1") +
    labs(title = "Unexposed",
         fill = "Perception of AI",
         x = "",
         y = "") +
    theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        # axis.title.x = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold")
    )

```

## Export

```{r export}

# Export table 1, alluvial plots, and crosstables (for alluvial plot figures)


# Alluvial plots as pdf:
ggsave(
    here("plots", "paper_alluvial.pdf"),
    plot = wrap_plots(alluv_users, alluv_nonusers, guides = "collect"),
    create.dir = TRUE,
    dpi = 600,
    device = "pdf",
    width = 20,
    height = 15,
    unit = "cm"
)
ggsave(
    here("plots", "paper_alluvial_weighted.pdf"),
    plot = wrap_plots(weighted_alluv_users, weighted_alluv_nonusers, guides = "collect"),
    create.dir = TRUE,
    dpi = 600,
    device = "pdf",
    width = 20,
    height = 15,
    unit = "cm"
)

## Crosstables:

# Unweighted
dput(x = unweighted_crosstabs,
     file = here("plots", "crosstab_unweighted.txt"))

# Weighted
dput(x = weighted_crosstabs,
     file = here("plots", "crosstab_weighted.txt"))

## Table 1:

openxlsx::write.xlsx(
    list(
        "Base table" = basetable,
        "Weighted basetable" = basetable_weighted,
        "By baseline perception" = basetable_by_baseline_perception,
        "crosstable_Users" = summary_crosstable_chatgpt,
        "crosstable_Non-users" = summary_crosstable_nonusers
    ),
    file = here("plots", "basetables.xlsx"),
    borders = "rows",
    colWidths = "auto"
)

```

## Run regression analyses and export estimates

```{r export-estimates}


# unweighted:
dput(
    x = dynamic_ref_regression_func(responses_weights, weights = NULL)[term == "chatgpt_useChatGPT use", .(
        baseline_perception,
        y.level,
        term,
        estimate,
        conf.low,
        conf.high,
        estimates_label,
        baseline_perception_empty,
        p.value
    )],
    file = here("plots", "coef_tables_unweighted.txt")
)

# target pop weighted:
dput(
    x = dynamic_ref_regression_func(responses_weights, weights = "weights_invitees_rescaled")[term == "chatgpt_useChatGPT use", .(
        baseline_perception,
        y.level,
        term,
        estimate,
        conf.low,
        conf.high,
        estimates_label,
        baseline_perception_empty,
        p.value
    )],
    file = here("plots", "coef_tables_weighted_invitees.txt")
)

# All estimates from main analysis - weighted to background population:
dput(
    x = dynamic_ref_regression_func(responses_weights, weights = "weights_background_rescaled")[, .(
        baseline_perception,
        y.level,
        term,
        estimate,
        conf.low,
        conf.high,
        estimates_label,
        baseline_perception_empty,
        p.value
    )],
    file = here("plots", "all_coefs_weighted_background.txt")
)
```

